---
title: "Pathway Based Clustering"
author: 
  - "Jaime Ñíguez"
jupyter:
  jupytext:
    formats: ipynb,Rmd  
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r message=FALSE, echo=TRUE}


set.seed(12345)
library(caret)
library(glmnet)
library(kableExtra)
library(ranger)
library(randomForest)
library(foreach)
library(doParallel)
library(MLmetrics)
library(rstatix)
library(dendextend)
library(pheatmap)
library(tidyr)
library(DT)
library(ggplot2)
library(plotly)
library(purrr)
library(clusterProfiler)
library(stringr)
library(org.Hs.eg.db)
library(circlize)
library(tibble)
library(msigdbr)
library(pheatmap)
library(networkD3)
library(ggpubr)
library(factoextra)
library(fpc)
library(cluster)
library(glmnet)
library(biomaRt)
library(gprofiler2)
library(FactoMineR)
library(e1071)
library(RColorBrewer)
library(patchwork) 
library(forcats)
library(fpc)
library(clValid)
library(clusterSim)
library(DESeq2)
library(ggrepel)
library(dplyr)



base_dir <- "/home/jaimeniguez"


```



# *Clustering and Pairwise Cluster Comparison

Based on the gene subsets obtained in the previous section, the following process will be performed:

- The transcriptomic data will be divided into k clusters using hierarchical clustering.

- All clusters will be compared against each other in a pairwise manner (1 vs 1) to identify significant differences in any of the progression-related covariates. Statistical tests will be applied to each pairwise comparison, and the resulting p-values will be adjusted using the Bonferroni method.

- For each significant difference, the direction of the difference will be assessed. If the difference indicates greater disease progression, a score of +1 will be assigned to the patients in the more severe cluster.

- A final score will be calculated for each patient, obtained by summing the scores resulting from all pairwise comparisons in which the patient belongs to the more severe cluster.


A dataframe will be generated with the following columns:

- Pathway name

- Cluster label and total number of clusters (k) for that iteration

- Number of samples (n) in the cluster

- Name of the significant covariate

- Patient sample ID

- Label of the cluster used as reference in the pairwise comparison

The clinical covariates used to determine patient severity are listed below (*Risk_Var*) and also are included covariates for which higher values indicate lower severity  in *Inv_Var*.

```{r}

# Grouped Tremor & PIGD
Risk_var <- c("mds_updrs_part_i_summary_score","code_upd2101_cognitive_impairment","code_upd2102_hallucinations_and_psychosis","code_upd2103_depressed_mood","code_upd2104_anxious_mood","code_upd2105_apathy","code_upd2106_dopamine_dysregulation_syndrome_features", "mds_updrs_part_ii_summary_score","mds_updrs_part_iii_summary_score","Tremor_Right_UM","Tremor_Left_UM","Tremor_all_Um","PIGD_UM", "ess_summary_score", "moca_total_score", "Abeta", "p-Tau", "Tau", "rbd_summary_score", "code_upd2hy_hoehn_and_yahr_stage", "apoe", "mapt", "AGE_ONSET", "mod_schwab_england_pct_adl_score")

Inv_Var <- c("moca_total_score", "mod_schwab_england_pct_adl_score", "AGE_ONSET")

```


## Severity Score Generation {#sec:score}

```{r}

dividing_data_cluster <- function(df_covariates, df_transcript, Subsets_pathways, n_clust) {
  
  for (pathway in names(Subsets_pathways)) {
    message("Pathway: ", pathway)
    valid_columns <- Subsets_pathways[[pathway]][Subsets_pathways[[pathway]] %in% colnames(df_transcript)]
    
    if (length(valid_columns) < 2) {
      warning("No valid columns found for pathway: ", pathway)
      next
    } 
    
    # Generar el subset transcriptómico
    df_transcript_pathway <- df_transcript[, valid_columns, drop = FALSE]
    
    # Eliminar genes constantes o con todos NA
    is_constant_or_na <- apply(df_transcript_pathway, 2, function(x) all(is.na(x)) || sd(x, na.rm = TRUE) == 0)
    df_transcript_pathway <- df_transcript_pathway[, !is_constant_or_na, drop = FALSE]
    
    # Si quedan menos de 2 genes, saltar el pathway
    if (ncol(df_transcript_pathway) < 2) {
      warning("Not enough valid genes after removing constant/NA genes for pathway: ", pathway)
      next
    }
    
    # Calcular matriz de correlación
    cor.pe <- cor(t(as.matrix(df_transcript_pathway)), method = "pearson")
    
    # Manejar correlaciones perfectas
    cor.pe[cor.pe >= 0.9999] <- 0.9999
    cor.pe[cor.pe <= -0.9999] <- -0.9999
    
    # Calcular matriz de distancias
    DissMatrix <- as.dist(1 - cor.pe)
    
    # Proceder con el clustering
    tryCatch({
      hclust_object <- hclust(DissMatrix, method = "ward.D2")
      cluster_assignments <- cutree(hclust_object, k = n_clust)
      df_covariates$cluster <- cluster_assignments
      
      full_cluster <- unique(df_covariates$cluster)
      
      for (i in 1:(length(full_cluster) - 1)) {
        for (j in (i + 1):length(full_cluster)) {
          
          cluster_i <- df_covariates[cluster_assignments == i, ] 
          cluster_j <- df_covariates[cluster_assignments == j, ]
          
          variables <- setdiff(colnames(cluster_i), "cluster")
          
          for (variable in variables) {
            
            # Si es numérica
            if (is.numeric(df_covariates[[variable]])) {
              
              num_test_result <- wilcox.test(cluster_i[[variable]], cluster_j[[variable]], 
                                             exact = min(length(cluster_i[[variable]]), length(cluster_j[[variable]])) < 20)
              
              num_test_result$p.value <- p.adjust(num_test_result$p.value, method = "bonferroni", 
                                                  n = n_clust * length(variables))
              
              if (!is.na(num_test_result$p.value) && num_test_result$p.value <= 0.05 && variable %in% Risk_var) {
                if (variable %in% Inv_Var) {
                  if (median(cluster_i[[variable]]) > median(cluster_j[[variable]])) {
                    sev_score_mod <- abs(median(cluster_i[[variable]]) - median(cluster_j[[variable]])) / 
                      max(median(cluster_i[[variable]]), median(cluster_j[[variable]]))
                    
                    row <- data.frame(
                      Pathway = rep(pathway, nrow(cluster_j)),
                      Cluster = rep(j, nrow(cluster_j)),
                      k = rep(n_clust, nrow(cluster_j)),
                      n = rep(nrow(cluster_j), nrow(cluster_j)),
                      Covariate = rep(variable, nrow(cluster_j)),
                      sample_id = rownames(cluster_j),
                      p_value = rep(num_test_result$p.value, nrow(cluster_j)),
                      VS = rep(i, nrow(cluster_j)),
                      SevScore = 1,
                      SevScore_mod = sev_score_mod
                    )
                    Sev_score_detail <- rbind(Sev_score_detail, row)
                  }
                } else {
                  sev_score_mod <- abs(median(cluster_i[[variable]]) - median(cluster_j[[variable]])) / 
                    max(median(cluster_i[[variable]]), median(cluster_j[[variable]]))
                  
                  row <- data.frame(
                    Pathway = rep(pathway, nrow(cluster_i)),
                    Cluster = rep(i, nrow(cluster_i)),
                    k = rep(n_clust, nrow(cluster_i)),
                    n = rep(nrow(cluster_i), nrow(cluster_i)),
                    Covariate = rep(variable, nrow(cluster_i)),
                    sample_id = rownames(cluster_i),
                    p_value = rep(num_test_result$p.value, nrow(cluster_i)),
                    VS = rep(j, nrow(cluster_i)),
                    SevScore = 1,
                    SevScore_mod = sev_score_mod
                  )
                  Sev_score_detail <- rbind(Sev_score_detail, row)
                }
              }
              
              # Si es factor
            } else if (is.factor(df_covariates[[variable]])) {
              full_classes <- union(levels(df_covariates[[variable]]), 
                                    unique(c(cluster_i[[variable]], cluster_j[[variable]])))
              full_classes <- sort(as.character(full_classes))
              
              for (upper_class in full_classes) {
                upper_numeric <- as.numeric(upper_class)
                lower_classes <- full_classes[as.numeric(full_classes) < upper_numeric]
                if (length(lower_classes) == 0) next
                
                numerador <- 0
                denominador <- upper_numeric - min(as.numeric(lower_classes))
                
                for (lower_class in lower_classes) {
                  cluster_i_name <- paste0("Cluster_", i)
                  cluster_j_name <- paste0("Cluster_", j)
                  
                  tab <- matrix(0, nrow = 2, ncol = 2,
                                dimnames = list(Clase = c(lower_class, upper_class),
                                                Grupo = c(cluster_i_name, cluster_j_name)))
                  tab[lower_class, cluster_i_name] <- sum(cluster_i[[variable]] == lower_class, na.rm = TRUE)
                  tab[upper_class, cluster_i_name] <- sum(cluster_i[[variable]] == upper_class, na.rm = TRUE)
                  tab[lower_class, cluster_j_name] <- sum(cluster_j[[variable]] == lower_class, na.rm = TRUE)
                  tab[upper_class, cluster_j_name] <- sum(cluster_j[[variable]] == upper_class, na.rm = TRUE)
                  
                  if (all(rowSums(tab) > 0)) {
                    test_to_use <- if (any(chisq.test(tab)$expected < 5)) fisher.test else chisq.test
                    result <- test_to_use(tab)
                    result$p.value <- p.adjust(result$p.value, method = "bonferroni", n = n_clust * length(variables))
                    
                    prop_upper_cluster_i <- tab[upper_class, cluster_i_name] / sum(tab[, cluster_i_name])
                    prop_upper_cluster_j <- tab[upper_class, cluster_j_name] / sum(tab[, cluster_j_name])
                    
                    if (result$p.value <= 0.05) {
                      if (prop_upper_cluster_i > prop_upper_cluster_j) {
                        numerador <- numerador + 1
                        target_cluster <- cluster_i
                        target_index <- i
                        target_vs <- j
                      } else if (prop_upper_cluster_j > prop_upper_cluster_i) {
                        numerador <- numerador + 1
                        target_cluster <- cluster_j
                        target_index <- j
                        target_vs <- i
                      }
                    }
                  }
                }
                
                if (numerador > 0 && denominador > 0) {
                  sev_score_mod <- numerador / denominador
                  row <- data.frame(
                    Pathway = rep(pathway, nrow(target_cluster)),
                    Cluster = rep(target_index, nrow(target_cluster)),
                    k = rep(n_clust, nrow(target_cluster)),
                    n = rep(nrow(target_cluster), nrow(target_cluster)),
                    Covariate = rep(variable, nrow(target_cluster)),
                    sample_id = rownames(target_cluster),
                    p_value = rep(NA, nrow(target_cluster)),
                    VS = rep(target_vs, nrow(target_cluster)),
                    SevScore = rep(1, nrow(target_cluster)),
                    SevScore_mod = rep(sev_score_mod, nrow(target_cluster))
                  )
                  Sev_score_detail <- rbind(Sev_score_detail, row)
                }
              }
            }
            
          }
        }
      }
      
    }, error = function(e) {
      warning("Error in pathway ", pathway, ": ", e$message)
    })
    
    assign("Sev_score_detail", Sev_score_detail, envir = .GlobalEnv)
  }
  
}
  
```



## Run Severity Score generation

```{r}

PPMI_label_M0 <- PPMI_label %>% filter(visit_name == "M0")
PPMI_label_M0$visit_name <- NULL
PPMI_exp_M0 <- PPMI_exp %>% filter(row.names(.) %in% rownames(PPMI_label_M0))


PPMI_label_M12 <- PPMI_label %>% filter(visit_name == "M12")
PPMI_label_M12$visit_name <- NULL
PPMI_exp_M12 <- PPMI_exp %>% filter(row.names(.) %in% rownames(PPMI_label_M12))


PPMI_label_M24 <- PPMI_label %>% filter(visit_name == "M24")
PPMI_label_M24$visit_name <- NULL
PPMI_exp_M24 <- PPMI_exp %>% filter(row.names(.) %in% rownames(PPMI_label_M24))


PPMI_label_M36 <- PPMI_label %>% filter(visit_name == "M36")
PPMI_label_M36$visit_name <- NULL
PPMI_exp_M36 <- PPMI_exp %>% filter(row.names(.) %in% rownames(PPMI_label_M36))

```



```{r eval=FALSE, warning=FALSE}

Sev_score_detail <- data.frame(
  Pathway = character(),
  Cluster = numeric(),
  k = numeric(),
  n = numeric(),
  Covariate = character(),
  sample_id = character(),
  p_value = numeric(),
  VS = numeric(),
  SevScore = numeric(),
  SevScore_mod = numeric(),
  avg_sil_width = numeric(),
  ch_index = numeric(),
  db_index = numeric() 
)


for (n in 2:14){ 
  (dividing_data_cluster(df_covariates = PPMI_label_M0, df_transcript = PPMI_exp_M0, Subsets_pathways = Subsets_pathways, n_clust = n))
}

Sev_score_detail_M0 <- Sev_score_detail

saveRDS(Sev_score_detail_M0, "Sev_score_detail_M0_noFilter_group_factor_vst_only_protein")



Sev_score_detail <- data.frame(
  Pathway = character(),
  Cluster = numeric(),
  k = numeric(),
  n = numeric(),
  Covariate = character(),
  sample_id = character(),
  p_value = numeric(),
  VS = numeric(),
  SevScore = numeric(),
  SevScore_mod = numeric(),
  avg_sil_width = numeric(),
  ch_index = numeric(),
  db_index = numeric()

)


for (n in 2:14){
  (dividing_data_cluster(df_covariates = PPMI_label_M12, df_transcript = PPMI_exp_M12, Subsets_pathways = Subsets_pathways, n_clust = n))
}

Sev_score_detail_M12 <- Sev_score_detail

saveRDS(Sev_score_detail_M12, "Sev_score_detail_M12_noFilter_group_factor_vst_only_protein")



Sev_score_detail <- data.frame(
  Pathway = character(),
  Cluster = numeric(),
  k = numeric(),
  n = numeric(),
  Covariate = character(),
  sample_id = character(),
  p_value = numeric(),
  VS = numeric(),
  SevScore = numeric(),
  SevScore_mod = numeric(),
  avg_sil_width = numeric(),
  ch_index = numeric(),
  db_index = numeric()

)


for (n in 2:14){
  (dividing_data_cluster(df_covariates = PPMI_label_M24, df_transcript = PPMI_exp_M24, Subsets_pathways = Subsets_pathways, n_clust = n))
}

Sev_score_detail_M24 <- Sev_score_detail

saveRDS(Sev_score_detail_M24, "Sev_score_detail_M24_noFilter_group_factor_vst_only_protein")




Sev_score_detail <- data.frame(
  Pathway = character(),
  Cluster = numeric(),
  k = numeric(),
  n = numeric(),
  Covariate = character(),
  sample_id = character(),
  p_value = numeric(),
  VS = numeric(),
  SevScore = numeric(),
  SevScore_mod = numeric(),
  avg_sil_width = numeric(),
  ch_index = numeric(),
  db_index = numeric() 

)


for (n in 2:14){
  (dividing_data_cluster(df_covariates = PPMI_label_M36, df_transcript = PPMI_exp_M36, Subsets_pathways = Subsets_pathways, n_clust = n))
}

Sev_score_detail_M36 <- Sev_score_detail


saveRDS(Sev_score_detail_M36, "Sev_score_detail_M36_noFilter_group_factor_vst_only_protein")


# saveRDS(Sev_score_detail_M0, "Sev_score_detail_M0_noFilter_group_factor_vst")
# saveRDS(Sev_score_detail_M12, "Sev_score_detail_M12_noFilter_group_factor_vst")
# saveRDS(Sev_score_detail_M24, "Sev_score_detail_M24_noFilter_group_factor_vst")
# saveRDS(Sev_score_detail_M36, "Sev_score_detail_M36_noFilter_group_factor_vst")


```





